Weather_covariance_matrices_flowering

### 2020-12-05 R/Analysis_v0.5_weather_covariance_Uhyp_flowering.Rmd
Add Mashhattan plotted by covariance matrix posterior weights exploration for FL50

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(bigsnpr)
library(mashr)
library(switchgrassGWAS)
library(rlist)
```
# TRIAL script
# -----------------------------
NB: Trial using covariance matrices derived from weather-related environmental cues as U_hyp in mash.

@@

1. Make dataframe of each env variable across all 8 sites
2. Compute correlations with pairwise complete observations
  2a. Compare correlation matrices - how dissimilar are they?
  2b. Compute correlations for Gulf, Midwest, & G&M subsets & compare
3. Experiment with changing within-site covariance to coefficient of variation for each site. Wouldn't expect these to be 1, really.

@@

### plot covariance matrices to look at their differences
```{r}
pvdiv_plot_cov <- function(covdf){
  U1 <- cor_phe
  if(isSymmetric(U1)){
        for(i in 1:nrow(U1)){
          for(j in 1:ncol(U1)){
            if(i < j){
              U1[i, j] <- NA
            }
          }
        }
  }
  U1 <- as_tibble(U1, rownames = "rowU", .name_repair = "unique") %>%
          pivot_longer(cols = -.data$rowU, names_to = "colU",
                       values_to = "covar") %>%
          filter(!is.na(.data$covar))
  U1$colU <- factor(U1$colU, levels = colnames(cor_phe))
  U1$rowU <- factor(U1$rowU, levels = colnames(cor_phe))

@@

```{r}
U_hyp <- list()
ggcov <- list()
phe_hyp <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl", "crain_gr2fl", "crain_1d", "crain_2d", "crain_3d", "crain_5d", "crain_7d")

@@

## Set up mash inputs

### mash G&M
```{r}
i=1 # site
j=1 # phe
k=3 # subpop: G&M, M, Gulf
for(k in c(3)){
phe_suffix <- paste("FL50", subpop_v[k], 
                    sep = "*.*")
snp <- snp_attach(inputfiles$SNPfiles[k])
gwas_rds <- pvdiv_results_in_folder(path = file.path(outputdir), 
                                      pattern = paste0("GWAS_datatable*.*",
                                                       phe_suffix))
phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -42) %>%
    str_replace(., "_$", "")
phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
out1phe <- pvdiv_bigsnp2mashr(path = file.path(outputdir), 
                           snp = snp, gwas_rds = gwas_rds, 
                           phenotypes = phenotypes, numSNPs = numSNPs, 
                           clump = TRUE, scaled = TRUE, saveoutput = TRUE,
                           suffix = paste0("FL50_", names(subpop_v)[k]))
}
```
Getting a lot of this: "error: chol(): decomposition failed"

## Run mash Gulf
```{r}
for(k in c(1)){
  phe_suffix <- paste("FL50", subpop_v[k], 
                      sep = "*.*")
  gwas_rds <- pvdiv_results_in_folder(path = file.path(outputdir), 
                                        pattern = paste0("GWAS_datatable*.*",
                                                         phe_suffix))
  phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -42) %>%
      str_replace(., "_$", "")
  phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
  numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
  m_out <- mash_standard_run(path = outputdir, numSNPs = numSNPs, 
                             U_hyp = U_hyp,
                             suffix = paste0("FL50_", names(subpop_v)[k]),
                             saveoutput = TRUE)
  }
```

@@

# 3. Analyse mash output

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(bigsnpr)
library(mashr)
library(switchgrassGWAS)
library(rlist)
library(rlang)
```
# TRIAL script
# -----------------------------
NB: Trial using covariance matrices derived from weather-related environmental cues as U_hyp in mash.

@@

1. Make dataframe of each env variable across all 8 sites
2. Compute correlations with pairwise complete observations
  2a. Compare correlation matrices - how dissimilar are they?
  2b. Compute correlations for Gulf, Midwest, & G&M subsets & compare
3. Experiment with changing within-site covariance to coefficient of variation for each site. Wouldn't expect these to be 1, really. Although mash does set the maximum of the diagonal to 1... hmm. I wonder why.

@@

## 1a. Covar plot function
Function to plot covariance matrices to look at their differences
```{r}
pvdiv_plot_cov <- function(covdf){
  U1 <- cor_phe
  if(isSymmetric(U1)){
        for(i in 1:nrow(U1)){
          for(j in 1:ncol(U1)){
            if(i < j){
              U1[i, j] <- NA
            }
          }
        }
  }
  U1 <- as_tibble(U1, rownames = "rowU", .name_repair = "unique") %>%
          pivot_longer(cols = -.data$rowU, names_to = "colU",
                       values_to = "covar") %>%
          filter(!is.na(.data$covar))
  U1$colU <- factor(U1$colU, levels = colnames(cor_phe))
  U1$rowU <- factor(U1$rowU, levels = colnames(cor_phe))

@@

## 1b. Make U_hyp
```{r}
U_hyp <- list()
ggcov <- list()
phe_hyp <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl", "crain_gr2fl", "crain_1d", "crain_2d", "crain_3d", "crain_5d", "crain_7d")

@@

## 2a. Set up mash inputs

### i. mash G&M
```{r}
i=1 # site
j=1 # phe
k=3 # subpop: G&M, M, Gulf
for(k in c(3)){
phe_suffix <- paste("FL50", subpop_v[k], 
                    sep = "*.*")
snp <- snp_attach(inputfiles$SNPfiles[k])
gwas_rds <- pvdiv_results_in_folder(path = file.path(outputdir), 
                                      pattern = paste0("GWAS_datatable*.*",
                                                       phe_suffix))
phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -42) %>%
    str_replace(., "_$", "")
phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
out1phe <- pvdiv_bigsnp2mashr(path = file.path(outputdir), 
                           snp = snp, gwas_rds = gwas_rds, 
                           phenotypes = phenotypes, numSNPs = numSNPs, 
                           clump = TRUE, scaled = TRUE, saveoutput = TRUE,
                           suffix = paste0("FL50_", names(subpop_v)[k]))
}
```
Getting a lot of this: "error: chol(): decomposition failed"

### ii. Run mash Gulf
```{r}
for(k in c(1)){
  phe_suffix <- paste("FL50", subpop_v[k], 
                      sep = "*.*")
  gwas_rds <- pvdiv_results_in_folder(path = file.path(outputdir), 
                                        pattern = paste0("GWAS_datatable*.*",
                                                         phe_suffix))
  phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -42) %>%
      str_replace(., "_$", "")
  phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
  numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
  m_out <- mash_standard_run(path = outputdir, numSNPs = numSNPs, 
                             U_hyp = U_hyp,
                             suffix = paste0("FL50_", names(subpop_v)[k]),
                             saveoutput = TRUE)
  }
```

@@

# 3. Analyse mash output
## -- analysis outline --
•	How many SNP effects load onto each covariance matrix?
•	Mashhattans colored by covariance matrices.
•	What patterns do the data-driven matrices capture that our weather matrices don’t capture?
•	We characterized patterns of differential sensitivity and antagonistic pleiotropy between all SNPs and all pairs of gardens.
•	We looked for enrichments of genes that could potentially affect flowering in the 20kb genomic windows surrounding the SNPs with significant effects on flowering.
•	Rerun greenup mash and rewrite greenup mash section.

## 3a. Covar Loadings
How many SNP effects load onto each covariance matrix?
```{r}
gulfm <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/gwas/BLUPs/single_sites/Strong_Effects19000SNPs_FL50_Gulf.rds")
midwm <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/gwas/BLUPs/single_sites/Strong_Effects33000SNPs_FL50_Midwest.rds")
gandm <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/gwas/BLUPs/single_sites/Strong_Effects19000SNPs_FL50_Gulf_and_Midwest.rds")
```

### i. Midwest
```{r}
m <- midwm
length(get_significant_results(m))  # 9907/63367
get_significant_results(m)[1:10]

covarm <- mash_plot_covar(m, saveoutput = TRUE)
get_U_by_mass(m)
mash_plot_Ulist(m, range = get_U_by_mass(m))
mash_plot_effects(m, n = 1, saveoutput = TRUE)
mash_plot_sig_by_condition(m, saveoutput = TRUE)
mash_plot_manhattan_by_condition(m, saveoutput = TRUE)

mash_out <- pvdiv_results_in_folder(outputdir, "Strong")
mash_plot_pairwise_sharing(effectRDS = file.path(outputdir, mash_out[3])) # sign
mash_plot_pairwise_sharing(effectRDS = file.path(outputdir, mash_out[6])) # sign & size
Midw_greenup_gxe <- get_GxE(m)

mash_plot_pairwise_sharing(corrmatrix = Midw_greenup_gxe$S_AP/Midw_greenup_gxe$S_all_pairwise, saveoutput = TRUE)
mash_plot_pairwise_sharing(corrmatrix = Midw_greenup_gxe$S_DS/Midw_greenup_gxe$S_all_pairwise, reorder = TRUE, saveoutput = TRUE)
mash_plot_pairwise_sharing(corrmatrix = Midw_greenup_gxe$S_2_no, reorder = FALSE)
mash_plot_pairwise_sharing(corrmatrix = Midw_greenup_gxe$S_all_pairwise, reorder = FALSE)

```

### ii. G&M
```{r}
length(get_significant_results(gandm))  # 24518/35735
get_significant_results(gandm)[1:10]

covargam <- mash_plot_covar(gandm, saveoutput = TRUE)
get_U_by_mass(gandm)
mash_plot_Ulist(gandm, range = get_U_by_mass(gandm), limits = FALSE)
```

### iii. Gulf
```{r}
length(get_significant_results(gulfm))  # 24518/35735
get_significant_results(gulfm)[1:10]

covarG <- mash_plot_covar(gulfm, saveoutput = TRUE)
get_U_by_mass(gulfm)
mash_plot_Ulist(gulfm, range = get_U_by_mass(gulfm))
```
```{r}
covargam$covar_df %>% 
  mutate(Type = case_when(grepl("ED_", `Covariance Matrix`) ~ "data-driven",
                          grepl("single_effect", `Covariance Matrix`) ~ "canonical",
                          grepl("simple_het", `Covariance Matrix`) ~ "canonical",
                          grepl("no_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("identity", `Covariance Matrix`) ~ "canonical",
                          grepl("equal_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("_and_Midwest$", `Covariance Matrix`) ~ "GAM_hyp",
                          grepl("Midwest$", `Covariance Matrix`) ~ "Midw_hyp",
                          grepl("Gulf$", `Covariance Matrix`) ~ "Gulf_hyp",
                          TRUE ~ "other"
                          )) %>%
  group_by(Type) %>%
  summarise(Mass = sum(Mass))

21.65+10.9795
covarG$ggobject
covarG$covar_df
8.38457+18.782188+18.38669
```

## 3b.	Mashhattans 
Mashhattans colored by covariance matrices.

### i. Function

I'm not sure that this function is really the way to go... I don't understand how SNPs really load onto these different covariance matrices.

Seems that the posterior weights for each SNP sum to one across this whole covariance grid. Sometimes there is a pw near 1, but more often they're a mix. So I think this is still useful for exploratory purposes - which covar have the max posterior weights? But I probably want to plot the posterior weights for each cov matrix one by one, which is much simpler to accomplish anyway.
```{r}
#' Return the name of the covariance matrix with the highest posterior weight
#'     for each effect
#'
#' @param m the mash result (from joint or 1by1 analysis)
#'
#' @return returns a df with Marker, Covariance_Matrix, and Grid_Value. I'm assuming that the grid value is an indication of like a svd decomp
#'
get_topcov = function(m){
  df <- switchgrassGWAS:::get_marker_df(m = m)
  covdf <- purrr::map_chr(df$value, ~ names(which(m$posterior_weights[.x,] ==
                                           max(m$posterior_weights[.x,])))) %>%
    enframe(name = "value", value = "Covariance_Grid") %>%
    separate("Covariance_Grid", into = c("Covariance_Matrix", "Grid_Value"),
             sep = "\\.", remove = FALSE) 
  df <- df %>% full_join(covdf, by = "value") %>% 
  mutate(Type = case_when(grepl("ED_", Covariance_Matrix) ~ "data-driven",
                          grepl("single_effect", Covariance_Matrix) ~ "canonical",
                          grepl("simple_het", Covariance_Matrix) ~ "canonical",
                          grepl("no_effects", Covariance_Matrix) ~ "none",
                          grepl("identity", Covariance_Matrix) ~ "canonical",
                          grepl("equal_effects", Covariance_Matrix) ~ "equal",
                          grepl("_and_Midwest$", Covariance_Matrix) ~ "GAM_hyp",
                          grepl("Midwest$", Covariance_Matrix) ~ "Midw_hyp",
                          grepl("Gulf$", Covariance_Matrix) ~ "Gulf_hyp",
                          grepl("_TX2$", Covariance_Matrix) ~ "canonical",
                          grepl("_TX1$", Covariance_Matrix) ~ "canonical",
                          grepl("_TX3$", Covariance_Matrix) ~ "canonical",
                          grepl("_M$", Covariance_Matrix) ~ "canonical",
                          grepl("_MO$", Covariance_Matrix) ~ "canonical",
                          grepl("_SD$", Covariance_Matrix) ~ "canonical",
                          grepl("_OK$", Covariance_Matrix) ~ "canonical",
                          grepl("_NE$", Covariance_Matrix) ~ "canonical",
                          grepl("_MI$", Covariance_Matrix) ~ "canonical",
                          TRUE ~ "other"
                          ))
  return(df)
}
```

### ii. Plot
```{r}
m <- midwm
log10bf_df <- switchgrassGWAS:::get_log10bf(m = m) %>%
    as.data.frame() %>%
    rownames_to_column(var = "value") %>%
    mutate(value = as.integer(.data$value)) %>%
    as_tibble() %>%
    left_join(switchgrassGWAS:::get_marker_df(m = m), by = "value") %>%
    dplyr::rename(log10BayesFactor = .data$V1) %>%
    dplyr::select(-.data$value)

#df <- get_topcov(m)

pw_df <- as_tibble(m$posterior_weights)
pw_df <- switchgrassGWAS:::get_marker_df(m = m) %>%
  bind_cols(pw_df)

ggman_df <- pw_df %>%
  #left_join(m_meta) %>%
  separate(.data$Marker, into = c("Chr", "Pos"), remove = FALSE, sep = "_",
           extra = "merge", convert = TRUE) %>%
  left_join(log10bf_df, by = "Marker") %>%
  arrange(.data$Chr, .data$Pos)

log10BF <- expression(paste("log"[10], plain("(Bayes Factor)")))

mash_plot_covar(m)
```

```{r}
gglist <- list()
for(i in seq_along(names(pw_df)[-(1:2)])){
var <- names(pw_df)[i+2]   # 20 or "Stand_BhatMidwest_FL50_M.21" is GIGO
gglist[[i]] <- ggman_df %>%
  #filter(Type != "canonical") %>%
  #group_by(crain_gr2fl_Gulf.8 > 0.01) %>% add_tally() %>%
  arrange(!! sym(var)) %>%
  ggplot(aes(x = .data$Pos, y = .data$log10BayesFactor)) +
  switchgrassGWAS::theme_oeco +
  geom_point(aes(color = !! sym(var), fill = !! sym(var),
                 shape = as.factor(.data$Chr))) +
  facet_wrap(~ .data$Chr, nrow = 1, scales = "free_x", strip.position = "bottom") +
  scale_color_viridis_c(option = "D", end = 0.95) + 
  scale_fill_viridis_c(option = "D", end = 0.95) +
  theme(strip.text = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.text.x = element_blank(),
        legend.text = element_text(size = 8),
        axis.ticks.x = element_blank(),
        panel.background = element_rect(fill=NA)) +
  labs(x = "Chromosome", y = log10BF) +
  scale_x_continuous(expand = c(0.32, 0.32)) +
  scale_shape_manual(values = rep(c(21,22),9), guide = FALSE) +
  geom_hline(yintercept = 2, linetype = 2) +
  ggtitle(var)
}
gglist
```
Midwest 4 cgdd gr2fl G&M background, everywhere, v low pw

G&M Mashhattans plotting SNP cov matrix posterior weights vs log10BF
4 "dyln_fl50_Gulf.3" lots of background
5 same lots of background
6 same background. Some above 2: Chr07N, Chr05K. 
7 same background. Some above 2 but all v low posterior weights.
8 background. Some above 2 Chr04N Chr09K.
9 PCA1 everywhere 
13 FL50 TX1 some very high log10BF and high posterior weights on many Chr.

Gulf Mashhattans by posterior weights for single covar mat/grids:
Gulf subpop covar matrix 11 (equal effects) has really interesting results - high posterior weights and high log10BF for several regions of the genome, esp end of Chr05N, end of Chr04K, Chr06N. Cool!
ED tPCA and ED PCA1 are of course everywhere. 19 & 17 especially.
NE one specific effect on Chr02N. (14, 15), maybe on Chr01N also.
SD (18) some specific high log10BF hits also: Chr03K, Chr09N, Chr09K, others.
TX1 (20) some specific high log10BF hits also: Chr07K, Chr05N, Chr09N, others.
TX2 (21) similar: Chr09N, Chr09N, Chr03N.

## 3c. Data-driven mat

What patterns do the data-driven matrices capture that our weather matrices don’t capture?

SVD and plots of effects for the ED_PCA matrices.

## 3d. DS & AP

We characterized patterns of differential sensitivity and antagonistic pleiotropy between all SNPs and all pairs of gardens.

## 3e. Enrichments
We looked for enrichments of genes that could potentially affect flowering in the 20kb genomic windows surrounding the SNPs with significant effects on flowering.

## 3f. Loglikelihood

log likelihood of models with and without U_hyp.

# -----------------------------
## In another script:
•	Rerun greenup mash and rewrite greenup mash section.

•	Rerun FL50 mash for Midwest without MI (or with reduced #'s for MI, OK, NE?). MI showing pop structure effects in mash results.
